---
title: "Chapter 2"
author: "Nick Lauerman"
output: 
  pdf_document: 
    fig_caption: yes
    fig_width: 4.5
    toc: yes
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries used
```{r Libraries}
library(Lahman)
```

## Data Sets
```{r Data Sets}

```


# Vectors
## Career of Wareen Spshn
```{r}
W <- c(8, 21, 15, 21, 21, 22, 14)
L <- c(5, 10, 12, 14, 17, 14, 19)
Win.Pct <- 100* W / (W +L)
Win.Pct
Year <- seq(1946, 1952)
Year
Year <- 1946:1952
Age <- Year - 1921
plot(Age, Win.Pct)
```

## Vector Functions
```{r}
mean(Win.Pct)
100 * sum(W) / (sum(W) + sum(L))
sort(W)
cumsum(W)
summary(Win.Pct)
```

## Vector index and logical variables
```{r}
W[c(1,2,5)]
W[1:4]
W[-c(1, 6)]
Win.Pct > 60
(W > 20) & (Win.Pct > 60)
Win.Pct == max(Win.Pct)
Year[Win.Pct == max(Win.Pct)]
Year[W + L > 30]
```

# Object and Container in R
## Character data nd matrices
```{r}
NL <- c("FLA",
        "STL",
        "HOU",
        "STL",
        "COL",
        "PHI",
        "PHI",
        "SFG",
        "STL",
        "SFG")
AL <- c("NYY",
        "BOS",
        "CHW",
        "DET",
        "BOS",
        "TBR",
        "NYY",
        "TEX",
        "TEX",
        "DET")
Winner <- c("NL",
            "AL",
            "AL",
            "NL",
            "NL",
            "NL",
            "AL",
            "NL",
            "NL",
            "NL")
N.Games <- c(6,
             4,
             4,
             5,
             4,
             5,
             6,
             5,
             7,
             4)
Year <- 2003:2012
results <- matrix(c(NL,AL), 10, 2)
results
dimnames(results)[[1]] <- Year
dimnames(results)[[2]] <- c("NL Teams", "AL Teams")
results
table(Winner)
barplot(table(Winner))
```

## Factors
```{r}

table(NL)
NL2 <- factor(NL, levels = c("FLA",
                             "PHI",
                             "HOU",
                             "STL",
                             "COL",
                             "SFG"))
str(NL2)
table(NL2)
```

## Lists
```{r}
World.Series <- list(Winner = Winner,
                     Number.Games = N.Games,
                     Seasons = "2003 to 2012")
World.Series$Number.Games
World.Series[[2]]
World.Series["Number.Games"]
```

# Collection of R Commands
## r scripts
```{r}
table(Winner)
barplot(table(Winner))
by(N.Games, Winner, summary)

```

## R Functions
```{r}
# also in scripts/Chapter2.R
hr.rates <- function(age, hr, ab){ 
   rates <- round(100 * hr / ab, 1)
   list(x=age, y=rates)
}
source(file = "scripts/Chapter2.R")
HR <- c(13,
        23,
        21,
        27,
        37,
        52,
        34,
        42,
        31,
        40,
        54)
AB <- c(341,
        549,
        461,
        543,
        517,
        533,
        474,
        519,
        541,
        527,
        514)
Age <- 19:29
hr.rates(Age, HR, AB)
plot(hr.rates(Age, HR, AB))
```

# Reading and Writing Data in R
## Importing data from a file
```{r}
spahn <- read.csv("./BookData/spahn.csv")
```

## Saving datasets
```{r}

HR
HR.Rates <- hr.rates(Age, HR, AB)
Mantle <- cbind(Age, HR, AB, Rates=HR.Rates$y)
write.csv(Mantle, "Mantle.csv", row.names = FALSE)
```

# Data Frames
## Introduction
```{r}
spahn[1:3, 1:10]
spahn[1,]
spahn[1 : 10, c("Age",
                "W",
                "L",
                "ERA")]
summary(spahn$ERA)
spahn$Age[spahn$ERA == min(spahn$ERA)]
```

## Manipulation with data frames
```{r}
spahn$FIP <- with(spahn, (13 * HR + 3 * BB - 2 * SO) / IP)
pos <- order(spahn$FIP)
head(spahn[pos, c("Year",
                  "Age",
                  "W",
                  "L",
                  "ERA",
                  "FIP")])
spahn1 <- subset(spahn,
                 Tm == "BSN" | Tm == "MLN")
by(spahn1[, c("W.L",
              "ERA",
              "WHIP",
              "FIP")],
   spahn1$Tm,
   summary)
```

## Merging and selectin from data frames
```{r}
NLBatting <- read.csv("BookData/NLbatting.csv")
ALBatting <- read.csv("BookData/ALbatting.csv")
batting <- rbind(NLBatting, ALBatting)
NLpitching <- read.csv("BookData/NLpitching.csv")
NL <- merge(NLBatting, NLpitching,
            by = "Tm")
NL.150 <- subset(NLBatting, HR > 150)

```

# Packages
```{r}
library(Lahman)
#?Batting
```










