---
title: "chapter 3"
author: "Nick Lauerman"
date: "Sunday, April 05, 2015"
output: html_document
---
# Chapter 3
RMarkdown of follow along work from chapter 3.
read in Hall of Fame (HOF) batting data set (no pitchers)
```{r read HOF batting, tidy=TRUE}
hof <- read.csv("~/R/Baseball/data/book/hofbatting.csv")

```

## a bar chart
 
```{r , tidy=FALSE}
# section 3.2.1
hof$MidCareer <- with(hof, (From + To) / 2)
hof$Era  <- cut(hof$MidCareer,
                breaks = c(1800, 1900, 1919, 1941, 1960, 1976, 1993, 2050),
                labels = c("19th Century",                           
                           "Dead Ball",
                           "Lively Ball",
                           "Intergration",
                           "Expansion",
                           "Free Agency",
                           "Long Ball"))
T.Era <- table(hof$Era)
T.Era
barplot(T.Era)

# section 3.2.2
barplot(table(hof$Era),
        xlab = "Era",
        ylab = "Frequency",
        main = "Era of the nonpitching Hall of Famers")

# section 3.2.3
plot(T.Era)
pie(T.Era)

```

## Saving graphs
```{r,tidy=FALSE}
# NOT INCLUDED
```

## Dot Plots
```{r, tidy=FALSE}
# section 3.4
dotchart(as.numeric(T.Era),
         labels = names(T.Era),
         xlab = "Frequency")

hof.500 <- subset(hof, HR >= 500)
hof.500 <- hof.500[order(hof.500$OPS), ]
dotchart(hof.500$OPS,
         labels = hof.500$X,
         xlab = "OPS")
```
## Stripchart and Histrogram
```{r, tidy=FALSE}
windows(width = 7, height = 3.5)
stripchart(hof$MidCareer,
           method = "jitter",
           pch = 1,
           xlab = "Mid Career")
hist(hof$MidCareer,
     xlab = "Mid Career",
     main="")
hist(hof$MidCareer,
     xlab = "Mid Career",
     main="",
     breaks = seq(1880, 2000, by = 20))
```

## 2 numeric variables
### Scatterplot
```{r, tidy=FALSE}
with(hof, plot(MidCareer, OPS))
with(hof,lines(lowess(MidCareer,
                      OPS,
                      f=0.3)))
with(hof, identify(MidCareer, OPS, X, n=4))
# Section 3.6.2
with(hof, plot(OBP, SLG))
with(hof, plot(OBP, SLG,
               xlim = c(0.25, 0.50),
               ylim = c(0.28, 0.75),
               pch = 19,
               xlab = "On-Base Percentage",
               ylab = "Slugging Percentage"))
with(hof, plot(OBP, SLG,
               xlim = c(0.25, 0.50),
               ylim = c(0.28, 0.75),
               pch = 19,
               xlab = "On-Base Percentage",
               ylab = "Slugging Percentage"))
curve(0.7 - x, add=TRUE)
curve(0.8 - x, add=TRUE)
curve(0.9 - x, add=TRUE)
curve(1.0 - x, add=TRUE)
text(0.27, 0.42, "OPS= 0.7")
text(0.27, 0.52, "OPS= 0.8")
text(0.27, 0.62, "OPS= 0.9")
text(0.27, 0.72, "OPS= 1.0")
with(hof, identify(OBP, SLG, X, n=6))
```

##Numeric and a Factor Variable
### parallel stripcharts
```{r, tidy=FALSE}
hof$HR.Rate <- with(hof, HR / AB)
stripchart(HR.Rate ~ Era , data = hof)

par(plt = c(0.2, 0.94, 0.145, 0.883))
stripchart(HR.Rate ~ Era ,
           data = hof,
           method = "jitter",
           pch = 1,
           las = 2)
```


### Parallel boxplots
```{r, tidy=FALSE}
par(plt = c(0.2, 0.94, 0.145, 0.883))
boxplot(HR.Rate ~ Era,
        data = hof,
        las = 2,
        horizontal = TRUE,
        xlab = "HR Rate")

```

## Comparing Ruth, Aaron, Bonds, and A-Rod
### Getting the data
```{r, tidy=TRUE}
# section 3.8.1
master <- read.csv("~/R/Baseball/data/lahman/Master.csv")
getinfo <- function(firstname, lastname){
     playerline <- subset(master,
                          nameFirst == firstname & nameLast == lastname)
     name.code <- as.character(playerline$playerID)
     birthyear <- playerline$birthYear
     birthmonth <- playerline$birthMonth
     birthday <- playerline$birthDay
     byear <- ifelse(birthmonth <= 6, birthyear, birthyear + 1)
     list(name.code=name.code, byear=byear)
}

ruth.info <- getinfo("Babe","Ruth")
aaron.info <- getinfo("Hank","Aaron")
bonds.info <- getinfo("Barry","Bonds")
arod.info <- getinfo("Alex","Rodriguez")

ruth.info

# section 3.8.2
batting <- read.csv("~/R/Baseball/data/lahman/Batting.csv")

ruth.data <- subset(batting, playerID == ruth.info$name.code)
ruth.data$Age <- ruth.data$yearID - ruth.info$byear

aaron.data <- subset(batting, playerID == aaron.info$name.code)
aaron.data$Age <- aaron.data$yearID - aaron.info$byear


bonds.data <- subset(batting, playerID == bonds.info$name.code)
bonds.data$Age <- bonds.data$yearID - bonds.info$byear


arod.data <- subset(batting, playerID == arod.info$name.code)
arod.data$Age <- arod.data$yearID - arod.info$byear

```
### Consstruction the graph
```{r, tidy=FALSE}
with(ruth.data, plot(Age, cumsum(HR),
                     type = "l",
                     lty = 3,
                     lwd = 2,
                     xlab = "Age",
                     ylab = "Career Home Runs",
                     xlim = c(18, 45),
                     ylim = c(0, 800)))

with(aaron.data, lines(Age, cumsum(HR),
                       lty = 2,
                       lwd = 2))
with(bonds.data, lines(Age, cumsum(HR),
                       lty = 1,
                       lwd = 2))
with(arod.data,  lines(Age, cumsum(HR),
                       lty = 4,
                       lwd = 2))
legend(20,700,
       legend = c("Bonds",
                  "Aaron",
                  "Ruth",
                  "ARod"),
       lty = 1:4,
       lwd = 2)
```

## The 1998 Home Run Race
### Getting the data
```{r, tidy=FALSE}
# section 3.9.1
source("~/R/Baseball/scripts/retrosheet.R")
#the below function is in the above file
#parse.retrosheet.php(1998)
# after this script was run the resulting file was moved and will not be run again.

data1998 <- read.csv("~/R/Baseball/data/retro/all1998.csv")
fields <- read.csv("~/R/Baseball/data/book/fields.csv")
names(data1998) <- fields[,"Header"]

retro.ids <- read.csv("~/R/Baseball/data/book/retrosheetIDs.csv")
sosa.id <- as.character(subset(retro.ids,
                               FIRST == "Sammy" & LAST == "Sosa")$ID)
mac.id <- as.character(subset(retro.ids,
                               FIRST == "Mark" & LAST == "McGwire")$ID)

sosa.data <- subset(data1998, BAT_ID == sosa.id)
mac.data <- subset(data1998, BAT_ID == mac.id)
```
### Extracting the data
```{r, tidy=FALSE}
#section 3.9.2 

createdata <- function(d){
     d$Date <- as.Date(substr(d$GAME_ID, 4, 11),
                       format = "%Y%m%d")
     d <- d[order(d$Date), ]
     d$HR <- ifelse(d$EVENT_CD == 23, 1, 0)
     d$cumHR <- cumsum(d$HR)
     d[, c("Date","cumHR")]
}

mac.hr <- createdata(mac.data)
sosa.hr <- createdata(sosa.data)
head(sosa.hr)
```
### Constructiong the plot
```{r, tidy=FALSE}
plot(mac.hr,
     type = "l",
     lwd = 2,
     ylab = "Home Runs in the Season")

lines(sosa.hr,
      lwd = 2,
      col = "grey")

abline(h=62, lty = 3)
text(10440, 65, "62")

legend(10440, 20,
       legend = c("McGwire (70)",
                  "Sosa (62)"),
       lwd = 2,
       col = c("black",
               "grey"))
```
















































































