---
title: "Chapter 9"
author: "Nick Lauerman"
output:
  html_document:
    number_sections: no
    toc: yes
  pdf_document:
    toc: yes
    toc_depth: 4
---

***

Graphs are included in a seperate code block, when needed, to allow caching of data within the 
markdown enviroment allowing to reduced time to execute and complile the page.

***

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE)
# change Drive letter for home (F:/) or work (E:/) locations
```
# setup
Libraries used:
```{r libraries}
# library(plotrix)
# library(ggplot2)
# library(lattice)
library(car)
library(plyr)
library(MASS)
```

Setting the seed for random numbers. The authors didn't set a seed or share 
it if they had set on in the book, this has rendered most to the results in the
simulations chapter non-reproducable. A seed is set here to for self consistancy,
allowing the results to be reproduced internally.

```{r set seed, cache=TRUE, cache.path="../cache/"}
set.seed(894619)
```



# Section 9.2
## Section 9.2.2
```{r section 9.2.2, cache=TRUE, cache.path="../cache/"}
data2011 <- read.csv("../data/retro/all2011.csv",
                     header = FALSE)
fields <- read.csv("../data/book/fields.csv")
names(data2011) <- fields[ ,"Header"]


data2011$HALF_INNING <- with(data2011,
                             paste(GAME_ID, INN_CT, BAT_HOME_ID))
data2011$RUNS.SCORED  <-  with(data2011,
                                (BAT_DEST_ID >3) +
                                     (RUN1_DEST_ID > 3) +
                                     (RUN2_DEST_ID > 3) +
                                     (RUN3_DEST_ID > 3))

get.state <- function(runner1, runner2, runner3, out){
     runners <- paste0(runner1, runner2, runner3)
     paste(runners, out)
}

RUNNER1 <- ifelse(as.character(data2011[, "BASE1_RUN_ID"]) == "", 0, 1)
RUNNER2 <- ifelse(as.character(data2011[, "BASE2_RUN_ID"]) == "", 0, 1)
RUNNER3 <- ifelse(as.character(data2011[, "BASE3_RUN_ID"]) == "", 0, 1)
data2011$STATE <- get.state(RUNNER1, RUNNER2, RUNNER3,  data2011$OUTS_CT)

NRUNNER1 <- with(data2011,
                 as.numeric(RUN1_DEST_ID ==1 |
                                 BAT_DEST_ID == 1))
NRUNNER2 <- with(data2011,
                 as.numeric(RUN1_DEST_ID == 2 |
                                 RUN1_DEST_ID == 2 |
                                 BAT_DEST_ID == 2))
NRUNNER3 <- with(data2011, 
                 as.numeric(RUN1_DEST_ID == 3 |
                                 RUN2_DEST_ID == 3 |
                                 RUN3_DEST_ID == 3 |
                                 BAT_DEST_ID == 3))
NOUTS <- with(data2011, OUTS_CT + EVENT_OUTS_CT)
data2011$NEW.STATE <- get.state(NRUNNER1, NRUNNER2, NRUNNER3, NOUTS)

data2011 <- subset(data2011, (STATE != NEW.STATE) | (RUNS.SCORED > 0))

require(plyr)
data.outs <- ddply(data2011, .(HALF_INNING),
                   summarize,
                   Outs.Inning = sum(EVENT_OUTS_CT))
data2011 <- merge(data2011, data.outs)

# NOTE: the second subset command is modified from the one given in the book to 
# align with the goals of the two subset commands given the the text preciding 
# the code fragment.
data2011C <- subset(data2011, Outs.Inning == 3)
data2011C <- subset(data2011C, BAT_EVENT_FL == TRUE)

require(car)
data2011C$NEW.STATE <- recode(data2011C$NEW.STATE,
                              "c('000 3', '100 3','010 3','001 3',
                              '110 3','101 3','011 3','111 3') = '3' ")
```

## Section 9.2.3
```{r section 9.2.3, cache=TRUE, cache.path="../cache/"}
T.matrix <- with(data2011C,
                 table(STATE, NEW.STATE))
P.matrix <- prop.table(T.matrix, 1)

P.matrix <- rbind(P.matrix, c(rep(0, 24), 1))

P1 <- round(P.matrix["000 0", ], 3)
data.frame(Prob=P1[P1 > 0])

P2 <- round(P.matrix["010 2",], 3)
data.frame(Prob = P2[P2 > 0])
```

### Section 9.2.4
```{r Section 9.2.4, cache=TRUE, cache.path="../cache/"}
count.runners.outs <- function(s){
     sum(as.numeric(strsplit(s,"")[[1]]),
         na.rm = TRUE)
}

runners.outs <- sapply(dimnames(T.matrix)[[1]],
                      count.runners.outs)[-25]
R <- outer(runners.outs + 1,
           runners.outs,
           FUN = "-")
dimnames(R)[[1]] <- dimnames(T.matrix)[[1]][-25]
dimnames(R)[[2]] <- dimnames(T.matrix)[[1]][-25]

R <- cbind(R, rep(0,24))

simulate.half.inning <- function(P, R, start = 1){
     s <- start
     path <- NULL
     runs <- 0
     while(s < 25){
          s.new <- sample(1:25, 1, prob = P[s, ])
          path <- c(path, s.new)
          runs <- runs + R[s, s.new]
          s  <- s.new
     }
     runs
}

RUNS <- replicate(10000,
                  simulate.half.inning(T.matrix, R))
table(RUNS)

sum(RUNS[RUNS >= 5]) / 10000

mean(RUNS)


RUNS.J <- function(j){
     mean(replicate(10000,
                    simulate.half.inning(T.matrix, R, j)))
}
Runs.Expectancy <- sapply(1:24, RUNS.J)
Runs.Expectancy <- t(round(matrix(Runs.Expectancy, 3, 8), 2))
dimnames(Runs.Expectancy)[[2]] <- c("0 Outs", "1 Outs","2 Outs")
dimnames(Runs.Expectancy)[[1]] <- c("000","001","010","011","100",
                                    "101","110","111")
Runs.Expectancy

Runs <- matrix(c(0.47, 0.25, 0.10, 1.45, 0.94, 0.32,
                 1.06, 0.65, 0.31, 1.93, 1.34, 0.54,
                 0.84, 0.50, 0.22, 1.75, 1.15, 0.49,
                 1.41, 0.87, 0.42, 2.17, 1.47, 0.76),
               8, 3,  byrow = TRUE)
Runs - Runs.Expectancy
```

## Section 9.2.5
```{r section 9.2.5, cache=TRUE, cache.path="../cache/"}
P.matrix.3 <- P.matrix %*% P.matrix %*% P.matrix

sorted.P <- sort(round(P.matrix.3["000 0", ], 3),
                 decreasing = TRUE)
head(data.frame(Prob = sorted.P))

Q <- P.matrix[-25, -25]
N <- solve(diag(rep(1, 24)) - Q)

N.0000 <- round(N["000 0", ], 2)
head(data.frame(N = N.0000))
sum(N.0000)
Length <- round(t(N %*% rep(1, 24)), 2)
data.frame(L= Length[1, 1:8])

```

## Section 9.2.6
```{r section 9.2.6, cache=TRUE, cache.path="../cache/"}
data2011C$HOME_TEAM_ID  <- with(data2011C,
                               substr(GAME_ID, 1, 3))
data2011C$BATTING.TEAM <- with(data2011C,
                               ifelse(BAT_HOME_ID == 0,
                                      as.character(AWAY_TEAM_ID),
                                      as.character(HOME_TEAM_ID)))

Team.T <- with(data2011C,
               table(BATTING.TEAM,
                     STATE,
                     NEW.STATE))
d.state <- subset(data2011C, STATE == "100 2")
Team.T.S  <- with(d.state, 
                  table(BATTING.TEAM,
                        NEW.STATE))
Team.T.S
```

# Section 9.3
## Section 9.3.2
```{r section 9.3.2, cache=TRUE, cache.path="../cache/"}
make.schedule <- function(teams, k){
     n.teams <- length(teams)
     Home <- rep(gl(n.teams,
                    n.teams,
                    length = n.teams^2,
                    labels = teams),
                 k)
     Visitor <- rep(gl(n.teams,
                       1,
                       length = n.teams^2,
                       labels = teams),
                    k)
     schedule <- data.frame(Home = Home,
                            Visitor = Visitor)
     subset(schedule, Home != Visitor)
     }

NL <- c("ATL",
        "CHN",
        "CIN",
        "HOU",
        "LAN",
        "NYN",
        "PHI",
        "PIT",
        "SFN",
        "SLN")
AL <- c("BAL",
        "BOS",
        "CAL",
        "CHA",
        "CLE",
        "DET",
        "MIN",
        "NYA",
        "OAK",
        "WS2")
teams <- c(NL,AL)
league <- c(rep(1,10),
            rep(2,10))
schedule <- rbind(make.schedule(NL, 9),
                  make.schedule(AL, 9))
```

## Section 9.3.3
```{r section 9.3.3, cache=TRUE, cache.path="../cache/"}
s.talent <- 0.20
talents <- rnorm(20, 0, s.talent)
TAL <- data.frame(Team = teams,
                  League = league,
                  Talent = talents)
SCH <- merge(schedule,
             TAL,
             by.x = "Home",
             by.y = "Team")
names(SCH)[4] <- "Talent.Home"
SCH <- merge(SCH,
             TAL,
             by.x = "Visitor",
             by.y = "Team")
names(SCH)[6] <- "Talent.Visitor"
SCH$prob.Home <- with(SCH,
                      exp(Talent.Home)/(exp(Talent.Home) + exp(Talent.Visitor)))
head(SCH)
```

## Section 9.3.4
```{r section 9.3.4, cache=TRUE, cache.path="../cache/"}
SCH$outcome <- with(SCH,
                    rbinom(nrow(SCH), 1, prob.Home))
SCH$winner <- with(SCH,
                   ifelse(outcome,
                          as.character(Home),
                          as.character(Visitor)))
head(SCH[ , c("Visitor",
              "Home",
              "prob.Home",
              "outcome",
              "winner")])

wins <- table(SCH$winner)
WIN <- data.frame(Team = names(wins),
                  Wins = as.numeric(wins))
RESULTS <- merge(TAL, WIN)
```

## Section 9.3.5
```{r section 9.3.5, cache=TRUE, cache.path="../cache/"}
win.league <- function(RR, league){
     wins <- RR$Wins * (RR$League == league)
     MAX <- max(wins)
     if(sum(wins == MAX) > 1){
          prob <- exp(RR$Talent) * (wins == MAX)
          outcome <- c(rmultinom(1, 1, prob))
          RR$Winner.Lg <- RR$Winner.Lg + outcome
          }
     if(sum(wins == MAX) == 1){
          RR$Winner.Lg <- RR$Winner.Lg + as.numeric(wins == MAX)
          }
     RR
     }
RESULTS$Winner.Lg <- 0
RESULTS$Winner.WS <- 0
for(j in 1:2){
     RESULTS <- win.league(RESULTS, j)
     }
teams <- (1:20)[RESULTS$Winner.Lg == 1]
outcome <- c(rmultinom(1, 7, exp(RESULTS$Talent)[teams]))
winner <- teams[1] * (diff(outcome) < 0) + teams[2] * (diff(outcome) > 0)
RESULTS$Winner.WS[winner] <- 1
```

## Section 9.3.6
```{r section 9.3.6, cache=TRUE, cache.path="../cache/"}
one.simulation.68 <- function(s.talent){
     make.schedule <- function(teams, k){
          n.teams <- length(teams)
          Home <- rep(gl(n.teams,
                         n.teams,
                         length = n.teams^2,
                         labels = teams),
                      k)
          Visitor <- rep(gl(n.teams,
                            1,
                            length = n.teams^2,
                            labels = teams),
                         k)
          schedule <- data.frame(Home = Home,
                                 Visitor = Visitor)
          subset(schedule, Home != Visitor)
          }
     
     NL <- c("ATL",
             "CHN",
             "CIN",
             "HOU",
             "LAN",
             "NYN",
             "PHI",
             "PIT",
             "SFN",
             "SLN")
     AL <- c("BAL",
             "BOS",
             "CAL",
             "CHA",
             "CLE",
             "DET",
             "MIN",
             "NYA",
             "OAK",
             "WS2")
     teams <- c(NL,AL)
     league <- c(rep(1,10),
                 rep(2,10))
     schedule <- rbind(make.schedule(NL, 9),
                       make.schedule(AL, 9))
     talents <- rnorm(20, 0, s.talent)
     TAL <- data.frame(Team = teams,
                       League = league,
                       Talent = talents)
     SCH <- merge(schedule,
                  TAL,
                  by.x = "Home",
                  by.y = "Team")
     names(SCH)[4] <- "Talent.Home"
     SCH <- merge(SCH,
                  TAL,
                  by.x = "Visitor",
                  by.y = "Team")
     names(SCH)[6] <- "Talent.Visitor"
     SCH$prob.Home <- with(SCH,
                           exp(Talent.Home)/(exp(Talent.Home) + exp(Talent.Visitor)))
     SCH$outcome <- with(SCH,
                         rbinom(nrow(SCH), 1, prob.Home))
     SCH$winner <- with(SCH,
                        ifelse(outcome,
                               as.character(Home),
                               as.character(Visitor)))
     head(SCH[ , c("Visitor",
                   "Home",
                   "prob.Home",
                   "outcome",
                   "winner")])
     
     wins <- table(SCH$winner)
     WIN <- data.frame(Team = names(wins),
                       Wins = as.numeric(wins))
     RESULTS <- merge(TAL, WIN)
     win.league <- function(RR, league){
          wins <- RR$Wins * (RR$League == league)
          MAX <- max(wins)
          if(sum(wins == MAX) > 1){
               prob <- exp(RR$Talent) * (wins == MAX)
               outcome <- c(rmultinom(1, 1, prob))
               RR$Winner.Lg <- RR$Winner.Lg + outcome
               }
          if(sum(wins == MAX) == 1){
               RR$Winner.Lg <- RR$Winner.Lg + as.numeric(wins == MAX)
               }
          RR
          }
     RESULTS$Winner.Lg <- 0
     RESULTS$Winner.WS <- 0
     for(j in 1:2){
          RESULTS <- win.league(RESULTS, j)
          }
     teams <- (1:20)[RESULTS$Winner.Lg == 1]
     outcome <- c(rmultinom(1, 7, exp(RESULTS$Talent)[teams]))
     winner <- teams[1] * (diff(outcome) < 0) + teams[2] * (diff(outcome) > 0)
     RESULTS$Winner.WS[winner] <- 1
     return(RESULTS)
}
RESULTS <- one.simulation.68(0.20)
RESULTS


display.standing <- function(RESULTS, league){
     Standings <- subset(RESULTS,
                         League == league,
                         select = c("Team",
                                    "Wins"))
     Standings$Losses <- 162 - Standings$Wins
     Standings[order(Standings$Wins, decreasing = TRUE), ]
}
cbind(display.standing(RESULTS, 1),
      display.standing(RESULTS, 2))
with(RESULTS, as.character(Team[Winner.Lg == 1]))
with(RESULTS, as.character(Team[Winner.WS == 1]))

```

## Section 9.3.7
```{r sectioi 9.3.7 A, cache=TRUE, cache.path="../cache/"}
Many.Results <- NULL
for(j in 1:1000){
     Many.Results <- rbind(Many.Results, one.simulation.68(0.20))
}
```
```{r section 9.3.7 plot 1}
with(Many.Results, smoothScatter(Talent, Wins))
```
```{r sectioi 9.3.7 B, cache=TRUE, cache.path="../cache/"}
Results.avg <- subset(Many.Results,
                      Talent > -0.05 & Talent < 0.05)
```
```{r section 9.3.7 plot 2}
hist(Results.avg$Wins)
require(MASS)
truehist(Results.avg$Wins)
```
```{r sectioi 9.3.7 C, cache=TRUE, cache.path="../cache/"}
fit1 <- glm(Winner.WS ~ Talent,
            data = Many.Results,
            family=binomial)
fit2 <- glm(Winner.Lg ~ Talent,
            data = Many.Results,
            family=binomial)
```
```{r section 9.3.7 plot 3}
b1 <- coef(fit1)
curve(exp(b1[1] + b1[2] * x) / (1 + exp(b1[1] + b1[2] * x)),
      -0.4, 0.4,
      xlab = "Talent",
      ylab = "Probability",
      lwd = 1,
      lty = 1,
      ylim = c(0, 1))

b2 <- coef(fit2)
curve(exp(b2[1] + b2[2] * x) / (1 + exp(b2[1] + b2[2] * x)),
      add = TRUE,
      lwd = 1,
      lty = 2)
legend(-0.2, 0.8,
       legend = c("Win Pennant",
                  "Win World Series"),
       lwd = 1,
       lty = c(1, 2))
