---
title: "Chapter 9"
author: "Nick Lauerman"
output: pdf_document
---

***

Graphs are included in a seperate code block, when needed, to allow caching of data within the 
markdown enviroment allowing to reduced time to execute and complile the page.

***
Change in section 9.2.2 leaving the second subset to reflect whatis in the books code verse 
what is in the description and using all2011.csv supplied by author

```{rglobal_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE)
# change Drive letter for home (F:/) or work (E:/) locations
```
## setup
```{r libraries}
library(plotrix)
library(ggplot2)
library(lattice)
library(car)
library(plyr)
```


## Section 9.2
### Section 9.2.2
```{r section 9.2.2, cache=TRUE, cache.path="E:/Baseball/cache/"}
data2011 <- read.csv("E:/Baseball/data/book/all2011.csv",
                     header = FALSE)
fields <- read.csv("E:/Baseball/data/book/fields.csv")
names(data2011) <- fields[ ,"Header"]


data2011$HALF_INNING <- with(data2011,
                             paste(GAME_ID, INN_CT, BAT_HOME_ID))
data2011$RUNS.SCORED  <-  with(data2011,
                                (BAT_DEST_ID >3) +
                                     (RUN1_DEST_ID > 3) +
                                     (RUN2_DEST_ID > 3) +
                                     (RUN3_DEST_ID > 3))

get.state <- function(runner1, runner2, runner3, out){
     runners <- paste0(runner1, runner2, runner3)
     paste(runners, out)
}

RUNNER1 <- ifelse(as.character(data2011[, "BASE1_RUN_ID"]) == "", 0, 1)
RUNNER2 <- ifelse(as.character(data2011[, "BASE2_RUN_ID"]) == "", 0, 1)
RUNNER3 <- ifelse(as.character(data2011[, "BASE3_RUN_ID"]) == "", 0, 1)
data2011$STATE <- get.state(RUNNER1, RUNNER2, RUNNER3,  data2011$OUTS_CT)

NRUNNER1 <- with(data2011,
                 as.numeric(RUN1_DEST_ID ==1 |
                                 BAT_DEST_ID == 1))
NRUNNER2 <- with(data2011,
                 as.numeric(RUN1_DEST_ID == 2 |
                                 RUN1_DEST_ID == 2 |
                                 BAT_DEST_ID == 2))
NRUNNER3 <- with(data2011, 
                 as.numeric(RUN1_DEST_ID == 3 |
                                 RUN2_DEST_ID == 3 |
                                 RUN3_DEST_ID == 3 |
                                 BAT_DEST_ID == 3))
NOUTS <- with(data2011, OUTS_CT + EVENT_OUTS_CT)
data2011$NEW.STATE <- get.state(NRUNNER1, NRUNNER2, NRUNNER3, NOUTS)

data2011 <- subset(data2011, (STATE != NEW.STATE) | (RUNS.SCORED > 0))

require(plyr)
data.outs <- ddply(data2011, .(HALF_INNING),
                   summarize,
                   Outs.Inning = sum(EVENT_OUTS_CT))
data2011 <- merge(data2011, data.outs)
data2011C <- subset(data2011, Outs.Inning == 3)
data2011C <- subset(data2011, BAT_EVENT_FL == TRUE)

require(car)
data2011C$NEW.STATE <- recode(data2011C$NEW.STATE,
                              "c('000 3', '100 3','010 3',
                              '001 3','110 3','101 3','011 3',
                              '111 3') = '3' ")
```
```{r}
table(data2011C$Outs.Inning)
```

### Section 9.2.3
```{r section 9.2.3, cache=TRUE, cache.path="E:/Baseball/cache/"}
T.matrix <- with(data2011C,
                 table(STATE, NEW.STATE))
P.matrix <- prop.table(T.matrix, 1)

P.matrix <- rbind(P.matrix, c(rep(0, 24), 1))

P1 <- round(P.matrix["000 0", ], 3)
data.frame(Prob=P1[P1 > 0])

P2 <- round(P.matrix["010 2",], 3)
data.frame(Prob = P2[P2 > 0])
```

### Section 9.2.4
```{r Section 9.2.4, cache=TRUE, cache.path="E:/Baseball/cache/"}
count.runners.outs <- function(s){
     sum(as.numeric(strsplit(s,"")[[1]]),
         na.rm = TRUE)
}

runners.outs <- sapply(dimnames(T.matrix)[[1]],
                      count.runners.outs)[-25]
R <- outer(runners.outs + 1,
           runners.outs,
           FUN = "-")
dimnames(R)[[1]] <- dimnames(T.matrix)[[1]][-25]
dimnames(R)[[2]] <- dimnames(T.matrix)[[1]][-25]

R <- cbind(R, rep(0,24))

simulate.half.inning <- function(P, R, start = 1){
     s <- start
     path <- NULL
     runs <- 0
     while(s < 25){
          s.new <- sample(1:25, 1, prob = P[s, ])
          path <- c(path, s.new)
          runs <- runs + R[s, s.new]
          s  <- s.new
     }
     runs
}

RUNS <- replicate(10000,
                  simulate.half.inning(T.matrix, R))
table(RUNS)

sum(RUNS[RUNS >= 5]) / 10000

mean(RUNS)


RUNS.J <- function(j){
     mean(replicate(10000,
                    simulate.half.inning(T.matrix, R, j)))
}
Runs.Expectancy <- sapply(1:24, RUNS.J)
Runs.Expectancy <- t(round(matrix(Runs.Expectancy, 3, 8), 2))
dimnames(Runs.Expectancy)[[2]] <- c("0 Outs", "1 Outs","2 Outs")
dimnames(Runs.Expectancy)[[1]] <- c("000","001","010","011","100",
                                    "101","110","111")
Runs.Expectancy

Runs <- matrix(c(0.47, 0.25, 0.10, 1.45, 0.94, 0.32,
                 1.06, 0.65, 0.31, 1.93, 1.34, 0.54,
                 0.84, 0.50, 0.22, 1.75, 1.15, 0.49,
                 1.41, 0.87, 0.42, 2.17, 1.47, 0.76),
               8, 3,  byrow = TRUE)
Runs - Runs.Expectancy
```

### Section 9.2.5
```{r, cache=TRUE, cache.path="E:/Baseball/cache/"}
P.matrix.3 <- P.matrix %*% P.matrix %*% P.matrix

sorted.P <- sort(round(P.matrix.3["000 0", ], 3),
                 decreasing = TRUE)
head(data.frame(Prob = sorted.P))

Q <- P.matrix[-25, -25]
N <- solve(diag(rep(1, 24)) - Q)

N.0000 <- round(N["000 0", ], 2)
head(data.frame(N = N.0000))
sum(N.0000)
Length <- round(t(N %*% rep(1, 24)), 2)
data.frame(L= Length[1, 1:8])

```

